### 2018-10-04 paper/complexity.tex
New example and extended proof of Theorem 5 in section on Complexity Considerations.

The following example, however, shows that the maximal length $pq$ is really
required when testing for refinement.
\begin{example}\label{ex:pq}
Consider the CSP reference process $P$ and an erroneous implementation $Q$
specified as follows.

\begin{center}
\begin{minipage}{.4\textwidth}
\begin{eqnarray*}
P & = & a \then P_1 \intchoice b \then P_1 \intchoice c \then P_1
\\
P_1 & = & a \then P \extchoice b\then P
\end{eqnarray*}
\end{minipage}
\hfill
\begin{minipage}{.4\textwidth}
\begin{eqnarray*}
Q & = & a\then Q_1 \extchoice b\then Q_1
\\
Q_1 & = & a\then Q_2 \extchoice b\then Q_2
\\
Q_2 & = & a\then Q \intchoice b\then Q
\end{eqnarray*}
\end{minipage}
\end{center}

\medskip
Obviously, $P$'s normalised transition graph has 2 nodes, while $Q$'s graph
has 3. It is easy to see (and can be checked with FDR4) that $P\lessdet_T Q$,
but $\neg(P\lessdet_F Q)$. Furthermore, it can also be shown using FDR4 that
the ``test passed condition''
\[
(\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(j))\hide \Sigma
\]
holds for $U_F(0),\dots,U_F(4)$, but fails for $U_F(5)$. So, the
non-conformance of $Q$ cannot be detected by any test trace of length less or
equal to 5, but is revealed (as expected from Theorem~\ref{th:failurestest})
by a trace of length 6, because the last event offered by the test $U_F(5)$
is refused by $Q$. \xbox
\end{example}
\fixme{But, there is no argument that the proofs based on graph products
corresponds to this process. In particular the hiding does not seem to play a
role.}
\fxnote{jp: should be clear now, since we have explained in Section 3 that only 
the graph product $G(P)\times G(Q)$ is   needed in the proofs}
%
Generalising Example~\ref{ex:pq}, it can be shown that for any pair
$2\le p,q \in\mathbb{N}$,
there exist reference processes with $p$ states
and implementation processes with $q$ states, such that
a violation of the trace refinement property
can only be detected with a trace of length $pq$. This is proven in the following
theorem.

\begin{theorem}\label{th:maxtracelen}
Let $2\le p,q \in\mathbb{N}$. Then there exists a reference process $P$ and an
implementation process $Q$ with the following properties.
\begin{enumerate}
\item $G(P)$ has $p$ states.
\item $G(Q)$ has $q$ states.
\item $P\not\lessdet_T Q$.
\item $\forall s\in\trc(Q): \#s < pq\implies s\in\trc(P)$.
\end{enumerate}
As a consequence, the upper bound $pq$ for the length of traces to be tested when checking for failures refinement cannot be reduced without losing the test suite's completeness property.
\end{theorem}
\begin{proof}
Given $2\le p,q \in\mathbb{N}$, define reference process $P$ and implementation process $Q$
over alphabet $\Sigma =\{ a,b\}$ as follows.
\begin{eqnarray*}
P & = &  P(0)
\\
P(k) & = & \big( a \then P(k) \big) \extchoice (k < p-1) \& \big( b \then P(k+1)\big)
\\
Q & = & Q(0)
\\
Q(k) & = & (k < q-1) \& \big( a \then Q(k+1)    \big) \extchoice
( k = q-1)\& \big( a\then Q(0) \extchoice b\then Q(0)  \big)
\end{eqnarray*}
\fixme{The internal choice became external. To test for traces refinement, yo
put forward just one adaptive tests. So, there is more going on here.I think
it is because your tests are checking both traces and failures refinement.}
Using regular expression notation, the traces of $P$ can be specified as
\[
\trc(P) = \prefs\big(  (a^*b)^{p-1}a^* \big),
\]
where $\prefs(M)$ denotes the set of all prefixes of traces in $M\subset\Sigma^*$,
including the traces of $M$ themselves.
The traces of $Q$ can be specified by
\[
\trc(Q) = \prefs\big( (a^{q-1}(a|b))^*  \big).
\]
It is easy to see that $\trc(Q)\not\subseteq\trc(P)$; for example, the trace
$(a^{q-1}b)^p$ is in $\trc(Q)\setminus\trc(P)$.

Let $s \in\trc(Q)$ be any trace of length $\#s = pq-1$. Then $s$ can be represented
by $s = (a^{q-1}(a|b))^{p-1}a^{q-1} \in \prefs\big( (a^{q-1}(a|b))^*  \big)$.
Then $s$ is also an element of $\trc(P)$, because $(a^{q-1}(a|b))^{p-1}a^{q-1}$
is also contained in $\prefs\big(  (a^*b)^{p-1}a^* \big)$: this is easy to see, since
$\prefs\big(  (a^*b)^{p-1}a^* \big)$ contains all finite sequences of $a$-events,
where at most $p-1$ events $b$ have been inserted.
\xbox
\end{proof}
%
Since Theorem~\ref{th:maxtracelen} just states that a violation of trace
refinement may remain undetected if only traces shorter than $pq$ are checked
during tests, it can also be applied to our trace refinement tests.
Therefore, test suites $\{ U_T(z) \}$ with $z<pq-1$ are not complete. It is
discussed in Section~\ref{sec:conc} how the number of test traces to be
executed by complete test suites for failures or trace refinement can still
be reduced {\it without} reducing the maximal length.

>>>>

The following example, however, shows that the maximal length $pq$ is really
required when testing for refinement.
%\begin{example}\label{ex:pq}
%Consider the CSP reference process $P$ and an erroneous implementation $Q$
%specified as follows.
%
%\begin{center}
%\begin{minipage}{.4\textwidth}
%\begin{eqnarray*}
%P & = & a \then P_1 \intchoice b \then P_1 \intchoice c \then P_1
%\\
%P_1 & = & a \then P \extchoice b\then P
%\end{eqnarray*}
%\end{minipage}
%\hfill
%\begin{minipage}{.4\textwidth}
%\begin{eqnarray*}
%Q & = & a\then Q_1 \extchoice b\then Q_1
%\\
%Q_1 & = & a\then Q_2 \extchoice b\then Q_2
%\\
%Q_2 & = & a\then Q \intchoice b\then Q
%\end{eqnarray*}
%\end{minipage}
%\end{center}
%
%
%\medskip
%Obviously, $P$'s normalised transition graph has 2 nodes, while $Q$'s graph
%has 3. It is easy to see (and can be checked with FDR4) that $P\lessdet_T Q$,
%but $\neg(P\lessdet_F Q)$. Furthermore, it can also be shown using FDR4 that
%the ``test passed condition''
%\[
%(\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(j))\hide \Sigma
%\]
%holds for $U_F(0),\dots,U_F(4)$, but fails for $U_F(5)$. So, the
%non-conformance of $Q$ cannot be detected by any test trace of length less or
%equal to 5, but is revealed (as expected from Theorem~\ref{th:failurestest})
%by a trace of length 6, because the last event offered by the test $U_F(5)$
%is refused by $Q$. \xbox
%\end{example}

\begin{figure}[htbp]
\begin{center}
\begin{minipage}{.4\textwidth}
\includegraphics[width=1.2\textwidth]{theorem5p.pdf}
\end{minipage}
\hfill
\begin{minipage}{.55\textwidth}
\includegraphics[width=1.2\textwidth]{theorem5q.pdf}
\end{minipage}
\caption{Transition graphs of $P$ (left) and $Q$ (right) from Example~\ref{ex:pq}
 for $p=3$ and $q=4$.}
\label{fig:examplepq}
\end{center}
\end{figure}

\begin{example}\label{ex:pq}
Consider 
\fxnote{jp: I have updated the example with $P$ and $Q$ as used in the proof of the theorem. This should make thinks more readable.}
\fixme{The internal choice became external. To test for traces refinement, yo
put forward just one adaptive tests. So, there is more going on here.I think
it is because your tests are checking both traces and failures refinement.}
\fxnote{jp: no -- you are right: we want {\it internal} choice for $P$. With external
choice, this only shows that when testing for trace refinement, the length $pq$ is required.}
the CSP reference process $P$ and an erroneous implementation $Q$
specified as follows.

\begin{eqnarray*}
P & = &  P(0)
\\
P(k) & = & (k < p-1) \& \big( (a \then P(k)) \intchoice ( b \then P(k+1))\big)
\\ & & \extchoice
\\ & & (k = p-1) \& (a \then P(k))
\\
Q & = & Q(0)
\\
Q(k) & = & (k < q-1) \& \big( a \then Q(k+1)    \big) 
\\ & & \extchoice
\\ & & ( k = q-1)\& \big( a\then Q(0) \extchoice b\then Q(0)  \big)
\end{eqnarray*}

\medskip
The normalised transition graphs of $P$ and $Q$ are depicted in Fig.~\ref{fig:examplepq}
for the case $p=3,\ q=4$. Using FDR4, it can be shown for concrete values of
$p$ and $q$ that the  ``test passed conditions''
\[
(\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(j))\hide \Sigma
\]
and
\[
(\epass\then\Stop) \lessdet_T (Q\parallel[\Sigma] U_T(j))\hide \Sigma
\]
hold for $j = 0,\dots,pq-2$. This means, that none of the test cases $U_F(j)$
and $U_T(j)$ are capable of detecting failures and trace refinement violations, 
respectively, if they only check traces up to length $pq-1$ (recall that this
corresponds to $j\le pq-2$).

Process $Q$, however, neither conforms to $P$ in the failures refinement 
relation, nor in the trace
refinement relation. This can only be seen when executing the test $U_F(pq-1)$ and
$U_T(pq-1)$, respectively. These test fail, so this shows that $P\not\lessdet_F Q$ and
$P\not\lessdet_F Q$ according to Theorem~\ref{th:failurestest} and Theorem~\ref{th:tracetest}. Moreover, this shows that
the maximal trace
length $pq$ to be investigated in the tests cannot be further reduced without losing
the completeness property of the test suites.
\xbox
\end{example}

\fixme{But, there is no argument that the proofs based on graph products
corresponds to this process. In particular the hiding does not seem to play a
role.}
\fxnote{jp: should be clear now, since we have explained in Section 3 that only 
the graph product $G(P)\times G(Q)$ is   needed in the proofs.}
%
Generalising Example~\ref{ex:pq}, it can be shown that for any pair
$2\le p,q \in\mathbb{N}$,
there exist reference processes $P$ with $p$ states
and implementation processes $Q$ with $q$ states, such that
a violation of the trace refinement property
can only be detected with a trace of length $pq$. This is proven in the following
theorem. In the proof, we use the processes $P$ and $Q$ introduced in Example~\ref{ex:pq}.

% ----------------------------------------------------------------------------
\begin{theorem}\label{th:maxtracelen}
Let $2\le p,q \in\mathbb{N}$. Then there exists a reference process $P$ and an
implementation process $Q$ with the following properties.
\begin{enumerate}
\item $G(P)$ has $p$ states.
\item $G(Q)$ has $q$ states.
\item $P\not\lessdet_T Q$, and therefore, also $P\not\lessdet_F Q$.
\item $\forall s\in\trc(Q): \#s < pq\implies s\in\trc(P)$.
\item $Q\ conf\ P$.
\end{enumerate}
As a consequence, the upper bound $pq$ for the length of traces to be tested when checking for failures refinement or trace refinement 
cannot be reduced without losing the test suite's completeness property.
\end{theorem}
% ----------------------------------------------------------------------------
\begin{proof}
Given $2\le p,q \in\mathbb{N}$, define reference process $P$ and implementation process $Q$ as in Example~\ref{ex:pq}. It is trivial to see that 
$G(P)$ has $p$ nodes and $G(Q)$ has $q$ nodes, so statements 1 and 2 of the theorem hold.

Using regular expression notation, the traces of $P$ can be specified as
\[
\trc(P) = \prefs\big(  (a^*b)^{p-1}a^* \big),
\]
where $\prefs(M)$ denotes the set of all prefixes of traces in $M\subseteq\Sigma^*$,
including the traces of $M$ themselves.
The traces of $Q$ can be specified by
\[
\trc(Q) = \prefs\big( (a^{q-1}(a|b))^*  \big).
\]
It is easy to see that $\trc(Q)\not\subseteq\trc(P)$; for example, the trace
$(a^{q-1}b)^p$ is in $\trc(Q)\setminus\trc(P)$, because $P$-traces may contain at most $p-1$ $b$-events. This proves statement~3 of the theorem.

Let $s \in\trc(Q)$ be any trace of length $\#s = pq-1$. Then $s$ can be represented
by $s = (a^{q-1}(a|b))^{p-1}a^{q-1} \in \prefs\big( (a^{q-1}(a|b))^*  \big)$.
Then $s$ is also an element of $\trc(P)$, because $(a^{q-1}(a|b))^{p-1}a^{q-1}$
is also contained in $\prefs\big(  (a^*b)^{p-1}a^* \big)$: this is easy to see, since
$\prefs\big(  (a^*b)^{p-1}a^* \big)$ contains all finite sequences of $a$-events,
where at most $p-1$ events $b$ have been inserted. This proves statement 4 of the theorem.

To prove statement~5, we observe that the specification of $P$ implies
\[
\minaccs(P/s) = \left\{
\begin{array}{ll}
\{ \{a\}, \{b\} \} & \text{for all $s\in\trc(P)$ with $\#s<p-1$.}
\\
\{ \{a\} \} & \text{for all $s\in\trc(P)$ with $\#s\ge p-1$.}
\end{array}
\right.
\]
and
\[
\minaccs(Q/s) = \left\{
\begin{array}{ll}
\{ \{a\}  \} & \text{for all $s\in\trc(Q)$ with $\#s \neq 0\mod(q-1)$.}
\\
\{ \{a,b\} \} & \text{for all $s\in\trc(P)$ with $\#s = 0\mod(q-1)$.}
\end{array}
\right.
\]
As a  consequence, the minimal acceptance set $A_P = \{a\}$ which is
contained in every $\minaccs(P/s)$ fulfils $A_P \subseteq A_Q$ for
any $A_Q\in\minaccs(Q/s)$, when $s\in\trc(P)\cap \trc(Q)$. Now Lemma~\ref{lemma:tgtrcref}, 
(\ref{eq:failrefb}) can be applied to conclude that $Q\ conf\ P$.
\xbox
\end{proof}
%
Since Theorem~\ref{th:maxtracelen} just states that a violation of trace
refinement may remain undetected if only traces shorter than $pq$ are checked
during tests, it can also be applied to our trace refinement tests.
Therefore, test suites $\{ U_T(z) \}$ with $z<pq-1$ are not complete. It is
discussed in Section~\ref{sec:conc} how the number of test traces to be
executed by complete test suites for failures or trace refinement can still
be reduced {\it without} reducing the maximal length.
